Text embedding models reveal high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-01-31 paper/main.tex
initial round of edits (incomplete)

\title{Text embedding models enable high-resolution insights into
conceptual knowledge and learning in classroom-like settings}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge in real-world educational contexts. Our approach embeds
course content in a high-dimensional
conceptual space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
viewed two lecture videos from the Khan Academy platform, interleaved
between three short multiple-choice quizzes. We applied our
framework to the lectures' transcripts and the text of the quiz questions to
quantify the conceptual content presented in each moment of video and knowledge probed by each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings demonstrate how a
small set of quiz questions may be used to obtain rich and meaningful
high-resolution insights into individuals' knowledge, and how it
changes over time as they learn.

@@

Suppose that a teacher had access to a complete ``map'' of everything a
student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well the
student knew the to-be-learned information already, or how much they knew about
related concepts. For some students, they could potentially optimize their
teaching efforts to maximize efficiency by focusing primarily on not-yet-known
content. For other students (or other content areas), it might be more
effective to optimize for direct connections between already known content and
new material. Observing how the student's knowledge changed over time,
in response to their teaching, could also help to guide the teacher towards
the most effective strategy for that individual student.

@@

We then created a pool of
multiple-choice quiz questions that would enable us to test participants'
knowledge about each individual lecture, as well as related content not
specifically presented in either video ( see Tab.~\questions). Participants answered
questions randomly drawn from each content area (lecture 1, lecture 2, and
general physics knowledge) on each of the three quizzes. Quiz 1 was intended to
assess participants' ``baseline'' knowledge before training, quiz 2 assessed knowledge
after watching the \textit{Four Fundamental Forces} video (i.e., lecture 1),
and quiz 3 assessed knowledge after watching the \textit{Birth of Stars} video
(i.e., lecture 2).

@@

\caption{\textbf{Modeling course content.} \textbf{A.
Building a document pool from sliding windows of text.} We decompose each
lecture's transcript into a series of overlapping sliding windows. The set of
transcript snippets (across all windows) may be treated as a set of
``documents'' for training a text embedding model. \textbf{B. Constructing lecture content \textit{trajectories}.}  After training our model on the sliding windows from both lectures, we transform each lecture into a ``trajectory'' through text embedding space by joining the embedding coordinates of successive
sliding windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
questions in a shared space.} We apply the same model (trained on the two lectures' windows) to both lectures, along with the text of each question in our pool (Tab.~\questions), to project them into a shared text embedding space. This results in one trajectory per lecture and one coordinate for each question. Here we have projected the 15-dimensional embeddings onto their first 3 principal components for visualization.}
\label{fig:sliding-windows}
\end{figure}

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work~\citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model~\citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model yielded a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics discovered by the model). Within this space, each lecture's sequence of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

\caption{\textbf{Lecture and question topic overlap.} \textbf{A. Topic 
weight variability}. The bar plots display the variance of each topic's weight 
across lecture timepoints (top row) and questions (bottom row); colors 
denote topics. The top-weighted words from the most ``expressive'' (i.e., 
variable across observations) topic from each lecture are displayed in the 
upper right (orange: topic 2; yellow-green: topic 4). The top-weighted words 
from the full set of topics may be found in Table~\topics. \textbf{B. Relationships between topic weight variability.} Pairwise correlations between the distributions of topic weight variance for each lecture and question category. Each row/column corresponds to a bar plot in panel A.}

@@

%This indicated that our model
%captures the conceptual contents of the lectures and quiz questions 
%sufficiently well to differentiate between questions relating to one lecture 
%versus the other, and could potentially be used to assess participants' knowledge for each lecture

% , and could be used to assess participants' knowledge about
%each lecture independently, based on the \textit{specific} questions they
%answered correctly or incorrectly on each quiz. 
%providing a slightly more 
%nuanced measure of what concepts they do and do not know than a simple 
%proportion-correct score.

%When we compared the
%variability in topic weights across each lecture's timepoints, and across
%questions from each category (i.e., lecture 1-related, lecture 2-related, and
%general physics knowledge), we found multiple interesting patterns (Fig.~\ref{fig:topics}). First,
%there is a strong correspondence between the sets of topics expressed by each
%lecture and its relevant questions. This indicates that our model is
%sensitive to the underlying conceptual content tested by different question,
%and should allow us to differentially assess participants' knowledge for the
%two lectures based on the \textit{specific} questions they answered correctly
%or incorrectly on each quiz. Second, the topics expressed by the general
%physics knowledge questions comprise a nonspecific mix of topics from the two
%lectures (4 from lecture 1; 5 from lecture 2), most of which are expressed to
%a relatively low degree. This indicates that our model should enable us to
%discriminate questions that are highly relevant to our assessment of
%participants' knowledge about concepts from the lectures from those that are
%not (or potentially less so). Third, the two lectures' trajectories vary
%along largely non-overlapping topic dimensions (e.g., variations in weight
%for topics 2, 3, 8, 12, 13, and 15 are specific to lecture 1; variations in
%weight for topics 1, 4, 5, 6, 7, and 10 are specific to lecture 2), but also
%share a small number of expressed topics in common (topics 11 and 14). This
%indicates that our lecture is able to simultaneously capture

This could enable us to estimate participants' knowledge for each of the two lectures, separately, by

This aspect of our model could enable us to assess participants' knowledge for each lecture separately, based on which \textit{specific} questions they answered right on wrong on each quiz, providing a slightly more refined estimate of the concepts they do and do not know than single proportion-correct score. However, while lectures are often organized around a single, broad theme at a coarse timescale, they typically cover...
%While individual lectures may be organized around a single broad theme at a coarse scale,

\textcolor{red}{\textbf{Paxton stopped here}}

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/lecture-question-similarity}
    \caption{\textbf{Which parts of each lecture are captured by each
    question?} Each panel displays timeseries plots showing how each question's
    topic vector correlates with each video timepoint's topic vector (Panel
    \textbf{A.}: correlations for the \textit{Four Fundamental Forces} lecture
    and associated questions; Panel \textbf{B.}: correlations for the
    \textit{Birth of Stars} lecture and associated questions). The colors denote
    question identities. The diamonds in each panel denote the moment of peak
    correlation between the indicated question and the lecture trajectory. The
    associated questions' text and snippets of the lectures' transcripts from the surrounding 30 seconds, are displayed at the bottom of the
    figure.}

@@

Although an individual lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

After fitting a topic model to each videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar (in terms of
Euclidean \textcolor{red}{should we say something like cosine instead of euclidean?} distance, correlation, etc.) topic vectors. In general, the
similarity between different documents' topic vectors may be used to
characterize the similarity in conceptual content between the documents.

@@

\begin{equation}
    \mathrm{ncorr}(x, y) = \frac{\mathrm{corr}(x, y) - \mathrm{mincorr}}{\mathrm{maxcorr} - \mathrm{mincorr}},
\end{equation}
and where $\mathrm{mincorr}$ and $\mathrm{maxcorr}$ are the minimum and maximum
correlations between any lecture timepoint and question, taken over all
timepoints and questions across both lectures and all five question used to estimate the knowledge trace \textcolor{red}{\textbf{Note: make sure this is correct in results section; not sure i caught all instances}}. We
also define $f(s, \Omega)$ as the $s$\textsuperscript{th} topic vector from the
set of topic vectors $\Omega$. Here $t$ indexes the set of lecture topic
vectors, $L$, and $i$ and $j$ index the topic vectors of questions used to estimate the knowledge trace, $Q$. Note that ``correct'' denotes the set of indices of
the questions the participant answered correctly on the given quiz.

>>>>

\title{Text embedding models reveal high resolution insights into conceptual
knowledge from short multiple choice quizzes}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural
language processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We applied our
framework to the videos' transcripts, and to text of the quiz questions, to
quantify the content of each moment of video and each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful high
resolution insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete tangible ``map'' of everything
a student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well
the student knew the to-be-learned information already, or how much they knew
about related concepts. For some students, they could potentially optimize
their teaching efforts to maximize efficiency by focusing primarily on
not-yet-known content. For other students (or other content areas), it might be
more effective to optimize for direct connections between already known content
and new material. Observing how the student's knowledge changed over time, in
response to their teaching, could also help to guide the teacher towards the
most effective strategy for that individual student.

@@

We created a pool of multiple-choice quiz questions that would enable us to
evaluate participants' knowledge about each individual lecture along with
related content not specifically presented in either video (see
Tab.~\questions). Participants answered questions randomly drawn from each
content area (lecture 1, lecture 2, and general physics knowledge) on each of
the three quizzes. Quiz 1 was intended to assess participants' ``baseline''
knowledge before training, quiz 2 assessed knowledge after watching the
\textit{Four Fundamental Forces} video (i.e., lecture 1), and quiz 3 assessed
knowledge after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

\caption{\textbf{Modeling course content.} \textbf{A. Building a document pool
from sliding windows of text.} We decompose each lecture's transcript into a
series of overlapping sliding windows. The set of transcript snippets (across
all windows) may be treated as a set of ``documents'' for training a text
embedding model. \textbf{B. Constructing lecture content
\textit{trajectories}.} After training our model on the sliding windows from
both lectures, we transform each lecture into a ``trajectory'' through text
embedding space by joining the embedding coordinates of successive sliding
windows parsed from its transcript. \textbf{C. Embedding multiple lectures and
questions in a shared space.} We apply the same model (trained on the two
lectures' windows) to both lectures, along with the text of each question in
our pool (Tab.~\questions), to project them into a shared text embedding space.
This results in one trajectory per lecture and one coordinate for each
question. Here we have projected the 15-dimensional embeddings onto their first
3 principal components for visualization.}

\label{fig:sliding-windows}
\end{figure}

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work~\citep{HeusEtal21} to extract the latent themes from the lectures'
contents using a topic model~\citep{BleiEtal03}. Briefly, topic models take as
input a collection of text documents and learn a set of ``topics'' (i.e.,
latent themes) from their contents. Once fit, a topic model can be used to
transform arbitrary (potentially new) documents into sets of ``topic
proportions,'' describing the weighted blend of learned topics reflected in
their texts. We parsed automatically generated transcripts of the two lectures
into overlapping sliding windows, each containing the text of the lecture
transcript from a particular time range. We treated the set of text (across all
of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's content. Intuitively, each window's ``topic vector'' (i.e.,
column of the topic-proportions matrix) is analogous to a coordinate in a
15-dimensional space whose axes are topics discovered by the model. Within this
space, each lecture's sequence of topic vectors (i.e., corresponding to sliding
windows parsed from its transcript) forms a \textit{trajectory} that captures
how its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B).
We resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

@@

\caption{\textbf{Lecture and question topic overlap.} \textbf{A. Topic weight
variability}. The bar plots display the variance of each topic's weight across
lecture timepoints (top row) and questions (bottom row); colors denote topics.
The top-weighted words from the most ``expressive'' (i.e., variable across
observations) topic from each lecture are displayed in the upper right (orange:
topic 2; yellow-green: topic 4). The top-weighted words from the full set of
topics may be found in Table~\topics. \textbf{B. Relationships between topic
weight variability.} Pairwise correlations between the distributions of topic
weight variance for each lecture and question category. Each row and column
corresponds to a bar plot in Panel A.}

@@

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

\begin{figure}[tp] \centering
\includegraphics[width=\textwidth]{figs/lecture-question-similarity}

\caption{\textbf{Which parts of each lecture are captured by each question?}
Each panel displays timeseries plots showing how each question's topic vector
correlates with each video timepoint's topic vector (Panel \textbf{A.}:
correlations for the \textit{Four Fundamental Forces} lecture and associated
questions; Panel \textbf{B.}: correlations for the \textit{Birth of Stars}
lecture and associated questions). The colors denote question identities. The
diamonds in each panel denote the moment of peak correlation between the
indicated question and the lecture trajectory. The associated questions' text
and snippets of the lectures' transcripts from the surrounding 30 seconds, are
displayed at the bottom of the figure.}

@@

@@

After fitting a topic model to each videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of Euclidean distance, correlation, or other geometric measures). In
general, the similarity between different documents' topic vectors may be used
to characterize the similarity in conceptual content between the documents.

@@

\begin{equation}
    \mathrm{ncorr}(x, y) = \frac{\mathrm{corr}(x, y) - \mathrm{mincorr}}{\mathrm{maxcorr} - \mathrm{mincorr}},
\end{equation}

and where $\mathrm{mincorr}$ and $\mathrm{maxcorr}$ are the minimum and maximum
correlations between any lecture timepoint and question, taken over all
timepoints and questions across both lectures and all five question used to
estimate the knowledge trace \textcolor{red}{\textbf{Note: make sure this is
correct in results section; not sure i caught all instances}}. We also define
$f(s, \Omega)$ as the $s$\textsuperscript{th} topic vector from the set of
topic vectors $\Omega$. Here $t$ indexes the set of lecture topic vectors, $L$,
and $i$ and $j$ index the topic vectors of questions used to estimate the
knowledge trace, $Q$. Note that ``correct'' denotes the set of indices of the
questions the participant answered correctly on the given quiz.
